---
title: "memisc vignette"
output: html_notebook
---
```{r}
knit_print.codebook <-function(x,...) 
  knitr::asis_output(format_html(x,...))

knit_print.descriptions <-function(x,...) 
  knitr::asis_output(format_html(x,...))

knit_print.ftable <-function(x,options,...)
  knitr::asis_output(
    format_html(x,
                digits=if(length(options$ftable.digits))
                          options$ftable.digits
                       else 0,
                ...))
# We can now adjust the number of digits after the comma
# for each column e.g. by adding an `ftable.digits` option
# to an R chunk, as in ```{r,ftable=c(2,2,0)}

knit_print.mtable <-function(x,...)
  knitr::asis_output(format_html(x,...))

```
```{r}
library(memisc)
options(digits=3)
nes1948 <-spss.portable.file("./nes1948.por")
print(nes1948)
names(nes1948)
```
```{r}
description(nes1948)
codebook(nes1948)
```
```{r}
vote.48 <- subset(nes1948,
                  select=c(
                    vote           = v480018,
                    occupation.hh  = v480029,
                    unionized.hh   = v480030,
                    gender         = v480045,
                    race           = v480046,
                    age            = v480047,
                    education      = v480048,
                    total.income   = v480049,
                    religious.pref = v480050
                  ))
str(vote.48)
```
```{r}
codebook(vote.48)
```
```{r}
vote.48 <- within(vote.48,{
  vote3 <- recode(vote,
    1 -> "Truman",
    2 -> "Dewey",
    3:4 -> "Other"
    )
  occup4 <- recode(occupation.hh,
    10:20 -> "Upper white collar",
    30 -> "Other white collar",
    40:70 -> "Blue collar",
    80 -> "Farmer"
    )
  relig3 <- recode(religious.pref,
    1 -> "Protestant",
    2 -> "Catholic",
    3:5 -> "Other,none"
    )
   race2 <- recode(race,
    1 -> "White",
    2 -> "Black"
    )
  })
ftable(xtabs(~vote3+occup4,data=vote.48))
```
May be better to use percentages.
```{r}
gt1 <- genTable(percent(vote3)~occup4,data=vote.48)
ftable(gt1,row.vars = 2)
```

Religion?
```{r}
gt2 <- genTable(percent(vote3)~relig3,data=vote.48)
ftable(gt2,row.vars=2)
```

Race?
```{r}
gt3 <- genTable(percent(vote3)~race2,data=vote.48)
ftable(gt3,row.vars=2)
```
Income?

```{r}
gt4 <- genTable(percent(vote3)~total.income,data=vote.48)
ftable(gt4,row.vars=2)
```

