---
title: "Protests slowed home price appreciation"
output: 
 pdf_document: 
    keep_tex: yes
    latex_engine: xelatex
---
# Were the protests
an inflection point for home prices?
In our meeting of Jan 1, 2017 we spent some time discussing this point,
I don't know how/if/where I can incorporate this into the proposal, but I
would like to share the results with you. As discussed, I find that prices
are still rising, but rising more slowly post-protest.
Visually, we graph home prices in all regions, as well as the national average, with a vertical black line at the June 2011 protests. There are
five years before the protests and 5 and 1/2 years post-protest. 
We are checking if the slope of the lines is indeed less-steep after the protests. We find that in the "Center and Jerusalem Periphery" and "Northern" regions, prices actually accelerated post-protest. In all other regions they still rose, but rose at a slower rate than they had before the protests.
Thus, the protests did demarcate an inflection point for housing prices. 
We then present these changes in tabular format.
```{r, cache=TRUE, include=FALSE}
library(censusFunctions)
library(ggplot2)
#library(pander)
library(scales)
source("./includes/getHousePrices.R", echo=FALSE)
source("./includes/importPrices.R",   echo=FALSE)
source("./includes/getGDP.R",         echo=FALSE)
source("./includes/graphGDPhouseholdConsumption.R",echo=FALSE)
source("./includes/groupHouseSize.R",echo=FALSE)

```
```{r,eval=FALSE, include=FALSE}
(p1.5$quarter)[1]  ## 2006
(p1.5$quarter)[47] ## 2016.5
(p1.5$quarter)[23]
     
     
```
```{r,include=FALSE}
require(ggplot2)
plot1.5rooms<-ggplot(data=p1.5,aes(x=quarter,
                     y=value,
                     group=variable
                     ))+geom_line(aes(colour=variable))+
                     ggtitle("Average P. 1.5-2.5 bedroom home:2006-2016q2")+theme(legend.title=element_blank(),
                   panel.grid.major.y = element_blank(),
                   panel.grid.minor = element_blank(),
                   axis.ticks.x = element_blank(),
                   axis.text.x=element_blank())+ ylab("NIS x 100k")+geom_vline(xintercept=28)+geom_vline(xintercept = 36)
```
```{r}
plot1.5rooms
ggsave(filename="P1.5.png", plot=plot1.5rooms)
```

Quarters 1-23, 2006q1 to 2011q2 are pre-protest. (5.5 years)
Quarters 24-47, 2011q3 to 2016q2 are post-protest. (5 years)

# Find the average yearly change in housing prices for each region in the 
pre- and post- protest environment.

```{r, cache=TRUE, include=FALSE}
require (scales)
## Haifa "h"
dh = Haifa1.5$value[23] - Haifa1.5$value[1]  ## 90.6 k NIS increase over T
ph = dh/Haifa1.5$value[1] ## 31% increase over T
yh = percent(ph / 5.5)        ## yearly increase
dh_post = Haifa1.5$value[47]-Haifa1.5$value[24]
ph_post = dh/Haifa1.5$value[24]
yh_post = percent(ph_post / 5)
yhN =ph/5.5
yhN_post=ph_post/5
## National Average "n"
dn = national1.5$value[23] - national1.5$value[1]
pn = dn/national1.5$value[1] 
yn = percent(pn / 5.5)
dn_post = national1.5$value[47]-national1.5$value[24]
pn_post = dn_post /national1.5$value[24]
yn_post = percent(ph_post / 5)
ynN = pn/5.5
ynN_post = pn_post/5
## Center "c"
dc = centerJeruPeri1.5$value[23] - centerJeruPeri1.5$value[1]  
pc = dh/centerJeruPeri1.5$value[1] 
yc = percent(pc / 5.5)        
dc_post = centerJeruPeri1.5$value[47]-centerJeruPeri1.5$value[24]
pc_post = dc/centerJeruPeri1.5$value[24]
yc_post = percent(pc_post / 5)
ycN = pc/5.5
ycN_post = pc_post/5
## Gush Dan "d"
dd = GushDan1.5$value[23] - GushDan1.5$value[1]  
pd = dh/GushDan1.5$value[1] 
yd = percent(pd / 5.5)        ## yearly increase
dd_post = GushDan1.5$value[47]-GushDan1.5$value[24]
pd_post = dh/GushDan1.5$value[24]
yd_post = percent(pd_post / 5)
ydN = pd/5.5
ydN_post = pd_post/5
## Jerusalem "j"
dj = Jerusalem1.5$value[23] - Jerusalem1.5$value[1]  
pj = dh/Jerusalem1.5$value[1] 
yj = percent(pj / 5.5)        ## yearly increase
dj_post = Jerusalem1.5$value[47]-Jerusalem1.5$value[24]
pj_post = dh/Jerusalem1.5$value[24]
yj_post = percent(pj_post / 5)
yjN = pj/5.5
yjN_post = pj_post/5
## South "s"
ds = South1.5$value[23] - South1.5$value[1] 
ps = dh/South1.5$value[1] 
ys = percent(ps / 5.5)        ## yearly increase
ds_post = South1.5$value[47]-South1.5$value[24]
ps_post = dh/South1.5$value[24]
ys_post = percent(ps_post / 5)
ysN = ps/5.5
ysN_post = ps_post/5
## North  "no"
dno = North1.5$value[23] - North1.5$value[1]  
pno = dno/North1.5$value[1] 
yno = percent(pno / 5.5)        ## yearly increase
dno_post = North1.5$value[47]-North1.5$value[24]
pno_post = dno_post/North1.5$value[24]
yno_post = percent(pn_post / 5)
ynoN = pno/5.5
ynoN_post = pno_post/5
## Haifa Suburbs    "q"
dq = qrayotHaifa1.5$value[23] - qrayotHaifa1.5$value[1]  
pq = dh/qrayotHaifa1.5$value[1] 
yq = percent(pq / 5.5)        ## yearly increase
dq_post = qrayotHaifa1.5$value[47]-qrayotHaifa1.5$value[24]
pq_post = dh/qrayotHaifa1.5$value[24]
yq_post = percent(pq_post / 5)
yqN = pq/5.5
yqN_post = pq_post/5
## Sharon "sh"
dsh = Sharon1.5$value[23] - Sharon1.5$value[1] 
psh = dh/Sharon1.5$value[1] 
ysh = percent(psh / 5.5)        ## yearly increase
dsh_post = Sharon1.5$value[47]-Sharon1.5$value[24]
psh_post = dh/Sharon1.5$value[24]
ysh_post = percent(psh_post / 5)
yshN = psh/5.5
yshN_post = psh_post/5
## Tel-Aviv "t"
dt = tlv1.5$value[23] - tlv1.5$value[1]
pt = dh/tlv1.5$value[1] 
yt = percent(pt / 5.5)        ## yearly increase before the protests
dt_post = tlv1.5$value[47]-tlv1.5$value[24]
pt_post = dh/tlv1.5$value[24]
yt_post = percent(pt_post / 5)
ytN = pt/5.5
ytN_post = pt_post/5
####
Regions = c("national","centerJeruPeri","GushDan","Haifa","Jerusalem",
               "South","North","qrayotHaifa","Sharon","tlv")
beforeProtests <-c(yn,yc,yd,yh,yj,ys,yno,yq,ysh,yt)
afterProtests <-c(yn_post,yc_post,yd_post,yh_post,
                  yj_post,ys_post,yno_post,yq_post,
                  ysh_post, yt_post)
change <-c("(4.35%)",
              "5.52%",
              ("(1.52%)"),
              "(1.13%)",
           "(0.86%)",
           "(2.83%)",
           "2.92%",
           "(0.55)%",
           "(0.35%)",
           "(0.87%)"
              )
m <- cbind(Regions,beforeProtests,afterProtests, change)
```

Most regions, and the national average experienced a slow-down
in housing price growth after the summer 2011 protests. Yearly average price appreciation is shown for homes of 1.5-2.5 rooms. The 
before protests period is q1 2006 to q2 2011.  The after protests period is q3 2011 to q2 2016.

The "Northern Region", and the "Center and Jerusalem Periphery
Region" went against this trend, price appreciation in these two regions actually accelerated post-protest. 
```{r}
m
```

